% Possible reviewers that I know personally include:
% - Chris Fields
% - Denis Baurain
% - Daisie Huang
% In addition, there are people that appear to be users but that I don't know:
% - Hayward, Alexander 
% - Jern, Patric
% The general idea here is that these are people I haven't published with (yet).
% The BioPerl people's page provides a murderer's row of other candidates, though
% quite a few I already have published with, so there's a potential for COI.
% Here's the list: https://github.com/orgs/bioperl/people
% Also, all people on the previous publication, or people that have since 
% been added to the Acknowledgements in the README.md are obviously out.



\documentclass{bioinfo}
\copyrightyear{2017} \pubyear{2017}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Applications Note}

% Application Notes (up to 2 pages; this is approx. 1300 words or 1000 words plus one figure)

\begin{document}
\firstpage{1}

\subtitle{Phylogenetics}

\title[short Title]{The Bio::Phylo libraries for phylogenetic data analysis, version 2.0}
\author[Sample \textit{et~al}.]{Rutger A. Vos\,$^{\text{\sfb 1,2}*}$ and Hannes Hettling\,$^{\text{\sfb 1}}$}
\address{$^{\text{\sf 1}}$Research group \textit{Endless Forms}, Naturalis Biodiversity Center, Leiden, P.O. Box 9517, 2300RA, The Netherlands and \\
$^{\text{\sf 2}}$Institute of Biology Leiden, Leiden University, Leiden, P.O. Box 9500, 2300RA, The Netherlands.}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{\textbf{Motivation:} Phylogenetic analysis is a broad and expanding field that
requires versatile programming toolkits to manage the various data types, file formats,
and needs for scalability, simulation, visualization, and data exploration.\\
\textbf{Results:} We present version 2.0 of the Bio::Phylo libraries for phylogenetic data
analysis. This new release represents a rewrite of the internal architecture, allowing
for extensions that improve speed and persistence, as well as greatly increased 
functionality in terms of analysis, data reading and writing, and visualization.\\
\textbf{Availability:} The package is released as open source software under the same terms as 
Perl itself and available from the comprehensive Perl archive network as well as directly
from the source code repository.\\
\textbf{Contact:} \href{rutger.vos@naturalis.nl}{rutger.vos@naturalis.nl}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

\maketitle

% RAV: here are some useful syntax tricks that we will need to observe:
% - refer to figures like so: 'Figure~\ref{fig:01}'
% - citations like this: '\citep{Bag01}' or like that: '\citet{Bag01}', citep is 
%   parenthisized, citet is in-text
% - equations as follows:
% \begin{equation}
% \sum \text{\it x}+ \text{\it y} =\text{\it Z}\label{eq:01}\vspace*{-10pt}
% \end{equation}
% which can then be referenced as Equation~(\ref{eq:01})

\section{Introduction}

% history
Phylogenetic data is multi-faceted, encompassing morphological observations, molecular 
sequences, biological taxa, and tree and network topologies, and is ubiquitous in a 
variety of research fields including comparative genomics, systematics, evolution, 
biodiversity research, and ecology. The types of operations that are performed on 
phylogenetic data range from data cleaning, integration, and conversions, to simulation,
character analysis and inference, and visualization. Flexible programming toolkits that
operate on phylogenetic data and aid in scripting these operations are therefore very 
useful and available in a variety programming languages. Bio::Phylo \citep{Vos2011} is 
the most versatile toolkit for handling phylogenetic data in the Perl programming
language.

% development process, including hackathons
Conceived about ten years ago \citep{Vos2006,Vos2017}, Bio::Phylo has been under
ongoing development ever since. Numerous people have contributed to it, by writing code 
- as volunteers and Google Summer of Code students - by submitting bug reports, and by 
providing inspiration and impetus in larger collaborative networks (e.g. 
\citet{Stoltzfus2010}, \citet{Koureas2016}, \citet{Koureas2016a}) and hackathons (e.g. 
\citet{Lapp2007}, \citet{Katayama2010}, \citet{Katayama2011}, \citet{Katayama2013}, 
\citet{Stoltzfus2013}, \citet{Katayama2014}, \citet{Vos2014}). In the process, additional
requirements surfaced, which have been addressed by a rewrite of the internal architecture
and the addition of functional modules. These have been integrated in a version 2.0
release, which we present here.

\section{Design}

% extensible design with XS subclasses
In `object-oriented' programming, the different concepts within the problem domain that
the software addresses are modelled as different classes (e.g. a class to model 
phylogenetic trees, and one to model multiple sequence alignments) that inherit some of 
their functionality from other such classes. For example, a class that represents the 
blueprint for a multiple sequence alignment might inherit certain generalized attributes 
of character state matrices (e.g. the number of characters and taxa) from a `super-class'. 
In turn, this multiple sequence alignment class can be inherited from to create a 
blueprint for a more specialized class, for example one that models amino-acid alignments. 

The object-oriented programming paradigm allows other programmers to extend the 
functionality of toolkits by inheriting from classes to address additional use cases, for 
example by improving performance in various ways. Bio::Phylo has always allowed for this, 
but the version 1.0 design made this relatively cumbersome because a lot of functionality 
would have to be re-implemented for any newly created class to integrate well in the rest 
of the toolkit. In the new design, each class has been reworked into two separate modules, 
one that performs the actual changes to any data associated with the class, and one that 
performs all the other user-friendly operations that do not directly affect the data. With 
this new design, a new class that only re-implements the changes to the data can re-use 
all the other operations, and thus be much more compact and easily written. In addition, 
the integration of such new classes has been made far more flexible (by adopting, in the 
parlance of software design patterns \citep{Gamma1995}, the `Factory' pattern throughout).

% object relational mapping

\section{Data input and output}

% file format

% megatrees

\section{Analysis and simulation algorithms}

% topology indices

% rankprob

\section{Visualization}

% unrooted trees

% radial trees

% clade labels

\section{Impact and re-use}

% studies using it, by type

% projects using it

\section{Availability}

All revisions of the source code are available from the source code repository at
http://github.com/rvosa/bio-phylo. The latest stable release version, which is usually 
considerably far behind the latest source code revision, is available from the 
Comprehensive Perl Archive Network (CPAN) at http://search.cpan.org/dist/Bio-Phylo. 
Accompanying this publication is a uniquely identifiable release, stamped with a Digital 
Object Identifier (DOI) issued by Zenodo.org: \texttt{doi:10.5281/zenodo.tbd}.

As is a common convention in Perl software releases, a dual licensing scheme applies to
Bio::Phylo - both the Artistic License 
(https://github.com/rvosa/bio-phylo/blob/master/COPYING) as well as the GNU General Public 
License (https://github.com/rvosa/bio-phylo/blob/master/LICENSE) applies. This is 
generally interpreted to mean that you are free to choose whichever of these licenses fits 
best with your own project, should you want to reuse all (or part) of Bio::Phylo. This is 
certainly the spirit: feel free to use these libraries however you see fit. No warranties.

\section*{Acknowledgements}

% Contributors on github
% earlier code contributors
% Mannis van Oven

\section*{Funding}

The research leading to these results has received funding from the European Community's 
Seventh Framework Programme (FP7/2007-2013) under grant agreement no. 237046.

\bibliographystyle{natbib}

\bibliography{document}

\end{document}
